Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-18 projectReport/chapter5.tex
M final edits to Ch.5 and Michelle edits up to Ch.4

From~\eqref{eqn_driving} we know that $g(u,x) = (1-a) x + a \tanh(Au + \alpha B x)$. Recall that $g$ is SI-invertible. We refer to $A$ as the input matrix and $B$ the reservoir matrix.

@@

We implement a feedforward network, (a network with no memory, unlike the RNN)\ednote{B: Expanded sentence s.t. FFN is in more context} required to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library \ednote{B: Is this something you want me to write in? M: We do not implement the RNN, but the FN that learns Gamma through the Tensorflow. We called the RNN + FN set up a RCN in Adriaan's paper. FYI: FN's work only if there is a definite relationship between the $u_n$ and $u_{n+1}$ which is rare in temporal data.
RNNs with ESP work when there is a definite relationship between $(\ldots,u_{n-1},u_n)$ and $u_{n+1}$ and is suited for temporal processing.}.

@@

Thirdly we prepare the network. Utilising a network $N$ neurons (the same value $N$ as in the dimensions of the matrices $A,B$). Learning of $\Gamma$ is done using the Adam Optimiser where the means square error (MSE) loss function is optimised. The system is trained several times (in our case 4) with incrementally smaller learning rates on each run (0.001 and thereafter divided by 10 at each run). 
Additional parameters such as the number of network-layers, neurons per layer and training-length are presented case-by-case below.

We then reach the prediction step. Making use of $g$'s SI-invertibility and $g$ as defined in~\eqref{eqn_driving}, we have the following expression for $u_n$ replicated from~\eqref{eqn_SI_RNN}.
\begin{equation*}
  u_n := A^{-1}\bigg(\overline{\tanh}^{^{-1}}\frac{1}{a}\Big(x_{n+1}-(1-a)x_n\Big) \bigg) - \alpha B x_n
\end{equation*}

The two model equations for learning from data~\eqref{eqns_from_data} allows us to determine what the state $x_{n+1}$ at time $n+1$ will be. 
We specifically remind ourselves of \[x_{k+1}=\pi_2 \circ (\Gamma, \pi_2) \circ (\pi_2,g) (u_k,x_k)\]

@@

We also consider a more general problem of learning an attractor of dynamical system when only observations of an orbit are given. Explicitly, if $(W,T)$ is a dynamical system with dynamics generated by $w_{n+1}=Tw_n$, and if $\theta:W \to \mathbb{R}$ is an observable, then the task is to learn a dynamical system that is topologically conjugate to $(W,T)$ and predict $\theta(w_{m+1}),\theta(w_{m+2}),\ldots$ using the data $\theta(w_{0}),\theta(w_{1}),\ldots,\theta(w_{m})$.  So how do we do it?
Suppose the input generated from the delay-coordinate map $\Phi_{\theta,2d}(\theta(w_{n})) := (\theta(w_{n-2d}),\ldots,\theta(w_{n-1}),\theta(w_{n}))$ is such that Takens delay embedding theorem (see Theorem~\ref{Thm_Takens}) holds, in which case, there exists a homeomorphism  $F_\theta: \Phi_{\theta,2d}(\theta(w_{n})) \mapsto \Phi_{\theta,2d}(\theta(w_{n+1}))$. Then if we feed  the input values $u_n := \Phi_{\theta,2d}(\theta(w_{n}))$ to a driven system $g$ that is SI-invertible and has USP, the induced system $(Y_F,G_F)$ would be topologically conjugate to the inverse-limit space of
$(\Phi_{\theta,2d}(W), F_\theta)$ as in Theorem~\ref{Thm_CET}, and hence one could forecast $u_n,u_{n+1},\ldots$. In other words, one could forecast  the values $\theta(w_n), \theta(w_{n+1}),\ldots$ by feeding these observations as delay coordinates. The advantage of feeding delay-coordinates to a driven system is that the embedding is stable in the sense of global dissipativity that we have described in the previous chapter, i.e., $Y_F$ is made to contained in an attractor of a mapping induced on $Y_F$ that is induced by $g$.

We note that the twin equations together~\ref{eqns_from_data} generate a model from data. For a chaotic system, a small amount of noise due either to the input-noise or computational error would cause the solutions from two accurate models with the same initial conditions to  diverge as they are not identical. \ednote{B: edited this sentence} However, statistically, they would be the same since due to ergodic theory all typical orbits have the same visiting frequency to any region of the phase space of the dynamical system. \ednote{B: I'll need to recap this sentence with you before presenting as it's new to me.} To illustrate that we plot the densities of the trajectories of the actual and predicted solutions.

Our implementation of the reservoir (A) and input (B) matrices in the RNN makes use of sparse matrices. We discuss three specifics relating to this concept. Firstly, the matrices are assigned a variable density in terms of the amount of non-zero entries and are scaled to have values within the range $[-1,1]$. For this, we make use of \textit{Python's SciPy} library and specifically the \textit{Sparse, LinAlg} modules. In our experiments, we generally opted for a density ratio of 10\%-25\%.
Secondly, we opted to enforce the requirement that both matrices be well-conditioned, i.e. that they have a condition number of \~1. The condition number of a matrix $M$ (denoted $cond(M)$) is the product of the matrix norms of the matrix $M$ and its inverse $M^{-1}$ and is a measure of the sensitivity of output of a matrix multiplication when an input is changed. \textbf{(Cite.)}\ednote{Ask M: Cite specific resource?} By ensuring a condition number close to 1, we also ensure that our driven function $g$, depending on the matrices $A$ and $B$, will not be sensitive to adjustments in the input. In so doing we then also guarantee another form stability. \ednote{B: Not complete enough. Edit further and cite}

Lastly, we ensure the matrix $B$ has unit spectral radius. This is necessary due to the Causal Embedding Theorem~\ref{Thm_CET} which can then guarantee the existence of semi-conjugacy between the systems $(Y_T, G_T)$ and $(\widehat{U}_T, \widehat{T})$. In fact, the theorem actually requires that $\alpha{B}$ have spectral radius $<1$, so we impose the restriction $\alpha<1$ and consider only the matrix $B$'s spectral radius from thereon.

Note that to learn $\Gamma: (x_{n-1},x_{n}) \mapsto u_n$ through our feedforward network, we could learn the principal components of $(x_{n-1},x_{n})$ as well. 
We employ this since it provides slightly better results empirically but is not essential.

Principal components are used only for a more efficient state representation to possibly reduce learning errors; it is not for a lossy approximation since we use all principal components \textbf{(Cite.)}\ednote{Ask M: Is there a specific resource to cite?}.
More explicitly, denote $X_{1:N}$ as the matrix with the first $N$ states of the network data represented by row vectors. If $X_{1:N}=U\Sigma P^T$\ednote{Expand here} represents the singular value decomposition of $X_{1:N}$ \ednote{Ask: Is there a specific resource to cite?}, we then denote the principal component matrix $P$, and the principal components are given by 
 $Z_{1:N}=X_{1:N}P$.
These principal components are then used to train a feedforward neural network. Denote the row vectors of $Z_{1:N}$ by 
$z_i^T, i=1,2,\ldots,N$ and the neural network by $NN$, the network is subsequently trained to learn an approximation of the map
$NN: (z_{n-1},z_n) \mapsto  u_n.
$
The learnt approximation of $NN$ can be used to approximate $\Gamma$ since 
\begin{equation}\label{Seq_RNN}
  NN\left( \begin{bmatrix} 
    P^Tx_{n-1} \\
    P^Tx_n
    \end{bmatrix}
    \right) = NN \circ 
    \begin{bmatrix}
    P^T & 0 \\
    0 & P^T 
    \end{bmatrix}\begin{bmatrix}
    x_{n-1}\\
    x_n
    \end{bmatrix} = u_n = \Gamma(x_{n-1},x_n).
\end{equation}

\section{Experimental Results}
In this section we showcase numerical results pertaining to 3 attractors that were simulated using the methodology described. We first consider a simple physical system that exhibits complicated dynamics.
\subsection{Double Pendulum}

The Double Pendulum consists of two pendulums fastened to one another such that the system moves as a whole. The first (or higher pendulum) is attached to a fixed point while the second is attached to the endpoint of the first. (See below the Figure~\ref{fig:dp_setup})
This system is a classical example of a system exhibiting chaos and is often used as explanation of the very well-known Butterfly Effect. \ednote{B: Originally though a reference here was needed, but no longer think so.} Consider again the Figure~\ref{fig:dp_sdic}.
The system has four variables: angle and angular velocity of each pendulum rod. An ideal system is assumed where the rods are massless and no friction in the pivots are considered.
\begin{figure}[ht]
  \includegraphics[scale=0.35]{Graphs/_dp_setup.eps}
  \centering
  \captionof{figure}{Pictorial representation of the setup of a DP. Image taken from~\cite{DPSetup}}
\label{fig:dp_setup}
\end{figure}

The scalar equations of motion are derived using Newtonian physics and are reproduced from~\cite{DPFormulas} in~\ref{eqns_dp}. Each pendulum component has an associated mass of the head, $m$, and length of the rod, $L_i$. Acceleration due to gravity is denoted by $g$ and initial velocities by $\dot{\theta}$
\begin{eqnarray}\label{eqns_dp}
  \ddot{\theta_{1}}  = \frac{-g(2m_1+m_2)\sin{\theta_1} - m_2g\sin(\theta_1-2\theta_2) - 2\sin(\theta_1-\theta_2)m_2({\dot{\theta_{2}}}^{2}L_2 + {\theta_{1}'}L_1\cos(\theta_1-\theta_2))} {L_1(2m_1 + m_2 -m_2\cos(2\theta_1 - 2\theta_2))}
  \\
  \ddot{\theta_{2}} = \frac{2\sin(\theta_1-\theta_2)(\theta_{1}'^{2}L_1(m_1+m_2) + g(m_1+m_2)\cos\theta_1 + \dot{\theta_{2}}^{2}L_2m_2\cos(\theta_1-\theta_2))}{L_2(2m_1 + m_2 -m_2\cos(2\theta_1 - 2\theta_2))}
\end{eqnarray}

The second-order system is converted to a first-order ODE system and solved using the Runge-Kutta numerical method.  
Here we chose the values $m_1=2.0$, $m_2=2.0$, $L_1=1.5$ and $m_2=1.5$ with initial velocities $v_1=1.5, v_2=-2$ for the respective pendulum heads.
Undamped DP data-sets were generated and used in this project due to the requirement that the map $T$ be surjective that we must not forget. 
An example of this would be the case of a DP exhibiting energy-loss (i.e. where the map $T$ is not surjective). In such a case, the attractor is a single point in the phase space when the double pendulum comes to rest. 
It must be mentioned once again that if we use data outside the attractor, learning is not reliable. 
To see this, consider Figure~\ref{fig:damped_pendulum} and~\ref{fig:dp_notsurjective}.

@@

\begin{figure}[ht]
  \includegraphics[scale=0.3]{Graphs/_dpfail_nonsurj.eps}
  \centering
\caption{Failure to forecast or learn a map which is not surjective: Evolution of the $y$-coordinate of the bottommost pendulum plotted here for 15000 timesteps. Plotted here the predicted value (red) versus the actual value (blue). The plots overlap for the first 10000 steps, the duration of training to learn the map $\Gamma$. Parameter-values are exactly as in the successful case discussed in the paragraph directly afterwards. Afterwards, the trajectories start diverging and it the predicted value no longer accurately represents the complexity or energy-loss of the actual data. }
\label{fig:dp_notsurjective}
\end{figure}
\ednote{Can I say in the caption \emph{Parameter-values are exactly as in the successful case discussed in the paragraph directly afterwards.}}

@@

The system was driven for 15500 steps, of which the first 500 were immediately discarded to simulate the network `forgetting'. The driving function $g$ had input and reservoir matrices were implemented as sparse matrices filled to 20\%. Parameter-values of $\alpha=0.99$ and $a=0.5$ were used.
The remaining 15000 datapoints were used to train a network with 128 neurons. 
The network was initialised to contain 16 hidden layers each with a layer dimension of 64, activated with the tanh activation function.
Training was accomplished using 256 epochs with batch sizes 128 and a delay of 1. See~\ref{fig:dp_success_traj} for results pertaining to the trajectory taken by the DP's components and then~\ref{fig:dp_success_density} to witness associated densities of predicted versus actual values.

@@

We now consider an experiment where noise is added to the system and find it to be incredibly robust. We employ the same parameters as before and add noise from a normal-distribution with mean zero and standard deviation equal to 0.1, translating to roughly 39dB's of noise.
(For the signal-to-noise ration we adopt the formula $\text{SNR}_{dB}=10\text{log}_{10}\big(\frac{P_\text{{signal}}}{P_{\text{noise}}}\big)$ where $P_\text{{signal}}$ refers to the variance observed in the normalised data and $P_{\text{noise}}$ represents the variance of the added noise.)

Once again consider the actual and predicted trajectories for the noisy DP dataset in Figure~\ref{fig:noisydp_success_traj}. Next we consider the densities of the predicted and actual values in Figure~\ref{fig:noisydp_success_density} and observe that they display highly similar distributions.

@@

 Lastly, a run was done where only a single coordinate (the $y$-coordinate of the lower pendulum head) was fed into the system and we can illustrate through the obtained success that the coordinate contains all the information in its past states necessary to forecast its future evolution.
 \textbf{Insert graphs here. Had to quickly change something here.}

@@

The behaviour of the system with a $\beta$-parameter value of 0.1056 was simulated by scaling it to fit inside the interval $[-1,1]$. Two cases were considered: noise-free and perturbed by noise. 
Noise was generated  by adding data points taken from a normal distribution with mean 0 and standard deviation 0.05, which to translates approximately 33dB. 
($\text{SNR}_{dB}$ is calculated exactly as for the DP-data above)

@@

Another 2-equation discrete-time map named the Fractal Dream Attractor, or more commonly known as the Pickover Map (first discovered by Clifford A. Pickover and discussed in his fascinating book `Chaos in Wonderland'\cite{PickoverChaos}), was also considered.
\begin{eqnarray}\label{eqns_clifford}
  {x_{n+1}=\sin(ay_n) + c\cos(ax_n)} \\
  {y_{n+1}=\sin(bx_n)+d\cos(by_n)}
\end{eqnarray}

@@

The behaviour of the system with a parameter values of $x_0=0$, $y_0=0$, $a = -1.7$,  $b = 1.8$, $c = -1.9$ and $d = -0.4$ was examined. 
Data was scaled to fit inside the interval $[-1,1]$. Only the noise-free case was investigated. Depictions of the empirical results available in Figure~\ref{fig:Clifford}.
 A sequence of 25000 observations was fed into the system with 15000 entries discarded to simulate the network's loss of memory. The next 1000 steps were predicted and compared with true data.
The RNN was initialised exactly as for Thomas' Cyclically Symmetric Strange attractor: the dimension of the network is set equal to 500 neurons and 12 hidden layers of dimension 64 each were initialised. Training is realised through 150 epochs of batch sizes of 128 each; the map is learnt via the Adam Optimiser using finer and finer learning rates (0.001 and then divided by 10 at every iteration).

\section{Reducing Functional Complexity in Noninvertible Maps}
\ednote{section header is overfull}
% Besides being able to learn the action of the Koopman operator exactly, our setup has other advantages over selecting or optimizing the choice of observables as in the Extended Dynamic Mode Decomposition (EDMD) algorithm (see, [21, 22]). We can alter the functional complexity of G T by changing the number of observables or equivalently by altering the dimension of X – in the case of a RNN implementation of g, by simply changing the number of neurons in the network. Empirically, increasing the dimension of X is found to increase the linear relationship (or intuitively reduces the functional complexity of G T ) that is measured as a generalization of the Pearson correlation coefficient to random vectors (see, [44]) between (x_{n−1} , x_{n}) and G_T (x_{n−1} , x_n), and such numerical evidence is tabulated in [11, Table 1]. In contrast, expanding the set of observables while using EDMD to learn a map with a lower functional complexity is a highly involved task, not least because so much that one often does not know how to guess a new observable, as well as the fact that adding observables does not necessarily retain the invariance of the span of the observables.

>>>>

From~\eqref{eqn_driving} we recall the equation of the RNN: $g(u,x) = (1-a) x + a \tanh(Au + \alpha B x)$. Recall that $g$ is SI-invertible when $A$ is invertible and has same dimension as $B$. We refer to $A$ as the input matrix and $B$ the reservoir matrix.

@@

We implement a feedforward network, (a network with no memory, unlike the RNN) required to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library 
% \ednote{B: Is this something you want me to write in? M: We do not implement the RNN, but the FFN that learns Gamma through Tensorflow. We called the RNN + FN set up a RCN in Adriaan's paper. FYI: FN's work only if there is a definite relationship between the $u_n$ and $u_{n+1}$ which is rare in temporal data. RNNs with ESP work when there is a definite relationship between $(\ldots,u_{n-1},u_n)$ and $u_{n+1}$ and is suited for temporal processing.}.

@@

Thirdly, we prepare the network by setting up a network with $N$ neurons (the same value $N$ as in the dimensions of the matrices $A,B$). Learning of $\Gamma$ is done using the Adam Optimiser where the means square error (MSE) loss function is optimised. 
The system is trained several times (in our case 4) with incrementally smaller learning rates on each run (0.001 and thereafter divided by 10 at each run). 
Additional parameters such as the number of network-layers, neurons per layer and training-length are presented case-by-case below.

Once $\Gamma$ is learnt then reach the prediction step, where we forecast $u_n$ from $(x_{n-1},x_{n})$. Note that  due to causal embedding $(x_{n-1},x_{n})$ is in one-to-one correspondence with the left-infinite orbit $(\ldots,u_{n-1},u_{n-1})$. Once $u_n$ is computed using $\Gamma$, we feed that to the driven system $g$ along with  $x_n$ to obtain $x_{n+1}$. Now we apply $\Gamma$ on $(x_n,x_{n+1})$ to forecast $u_{n+1}$ and the whole process repeats. Note that this corresponds to implementing the two model equations for learning from data~\eqref{eqns_from_data}.
% We then reach the prediction step. Making use of $g$'s SI-invertibility and $g$ as defined in~\eqref{eqn_driving}, we have the following expression for $u_n$ replicated from~\eqref{eqn_SI_RNN}.
% \begin{equation*}
%   u_n := A^{-1}\bigg(\overline{\tanh}^{^{-1}}\frac{1}{a}\Big(x_{n+1}-(1-a)x_n\Big) \bigg) - \alpha B x_n
% \end{equation*}

% The two model equations for learning from data~\eqref{eqns_from_data} allows us to determine what the state $x_{n+1}$ at time $n+1$ will be. 
% We specifically remind ourselves of \[x_{k+1}=\pi_2 \circ (\Gamma, \pi_2) \circ (\pi_2,g) (u_k,x_k)\]

@@

We also consider a more general problem of learning an attractor of dynamical system when only observations of an orbit are given. Explicitly, if $(W,T)$ is a dynamical system with dynamics generated by $w_{n+1}=Tw_n$, and if $\theta:W \to \mathbb{R}$ is an observable, then the task is to learn a dynamical system that is topologically conjugate to $(W,T)$ and predict $\theta(w_{m+1}),\theta(w_{m+2}),\ldots$ using the data $\theta(w_{0}),\theta(w_{1}),\ldots,\theta(w_{m})$.  So how do we do it?
Suppose the input generated from the delay-coordinate map $\Phi_{\theta,2d}(\theta(w_{n})) := (\theta(w_{n-2d}),\ldots,\theta(w_{n-1}),\theta(w_{n}))$ is such that Takens delay embedding theorem (see Theorem~\ref{Thm_Takens}) holds, in which case, there exists a homeomorphism  $F_\theta: \Phi_{\theta,2d}(\theta(w_{n})) \mapsto \Phi_{\theta,2d}(\theta(w_{n+1}))$. Then if we feed  the input values $u_n := \Phi_{\theta,2d}(\theta(w_{n}))$ to a driven system $g$ that is SI-invertible and has USP, the induced system $(Y_F,G_F)$ would be topologically conjugate to the inverse-limit space of
$(\Phi_{\theta,2d}(W), F_\theta)$ due to Theorem~\ref{Thm_CET}, and hence one could forecast $u_n,u_{n+1},\ldots$. In summary, one could forecast  the values $\theta(w_n), \theta(w_{n+1}),\ldots$ by feeding these observations as delay coordinates since the input is again an orbit from a dynamical system. 
The advantage of feeding delay-coordinates to a driven system  over learning a map that describes the evolution of the delay coordinates is that the embedding is stable in the sense of global dissipativity that we have described in the previous chapter, i.e., there is an attractor $A$ of a map that implements $G_T$ on an expansion $Y_F^+$ providing some robustness to noise.

Now discuss on another practical issue with forecasting chaotic dynamical systems.  
We note that the twin equations~\ref{eqns_from_data} generate a model from data. For a chaotic system, a small amount of noise due either to the input-noise or computational error would cause the solutions from even an accurate model with the same initial conditions to diverge. However, statistically, a chaotic system would behave well in the sense that the visiting frequency or the average time spent by an orbit in any region of the phase space of the dynamical system is identical for almost all initial conditions (almost all is specified as a measure), a result in ergodic theory (\cite{wikiErgodicTheory}) a topic beyond the scope of this thesis.
To illustrate the accuracy of the twin  equations~\ref{eqns_from_data} in modeling the data we plot the densities of the trajectories of the actual and predicted solutions in~\ref{noisydp_success_density, fig:dp_success_density}.

In our implementation, the input and reservoir  matrices $A$ and $B$ in the RNN are sparse matrices. This sparsity is not essential but seems to perform slightly empirically better than choosing the matrices arbitrarily.  We also discuss three specifics relating to this concept. Firstly, the matrices are assigned a variable density in terms of the amount of non-zero entries and are scaled to have values within the range $[-1,1]$. For this, we make use of \textit{Python's SciPy} library and specifically the \textit{Sparse, LinAlg} modules. In our experiments, we generally opted for a density ratio of 10\%-25\%.
Secondly, we opt to enforce the requirement that both matrices be well-conditioned, i.e. that they have a condition number of \~1. The condition number of a matrix $M$ (denoted $cond(M)$) is the product of the matrix norms of the matrix $M$ and its inverse $M^{-1}$ and is a measure of the sensitivity of output of a matrix multiplication when an input is changed.  We do not have any theoretical justifications for this.

%By ensuring a condition number close to 1, we also ensure that our driven function $g$, depending on the matrices $A$ and $B$, will not be sensitive to adjustments in the input.  In so doing we then also guarantee another form stability. \ednote{B: Not complete enough. Edit further and cite M: Please feel free to cite a wikipedia page}

% Lastly, we ensure the matrix $B$ has unit spectral radius. This is necessary due to the Causal Embedding Theorem~\ref{Thm_CET} which can then guarantee the existence of semi-conjugacy between the systems $(Y_T, G_T)$ and $(\widehat{U}_T, \widehat{T})$. In fact, the theorem actually requires that $\alpha{B}$ have spectral radius $<1$, so we impose the restriction $\alpha<1$ and consider only the matrix $B$'s spectral radius from thereon.
Lastly, we ensure the matrix $B$ has unit spectral radius so that the variable $\alpha$ in the RNN equation \eqref{eqn_driving} can be used to know if the USP is satisfied. It is a  feature that a matrix with a spectral radius $<1$ generally has the USP \cite{yildiz2012re}.

We remark on one more aspect that helps produce slightly better empirical results.  To learn $\Gamma: (x_{n-1},x_{n}) \mapsto u_n$ through our feedforward network, we could feed the principal components. 
Principal components are used only for a more efficient state representation to possibly reduce learning errors; it is not for a lossy approximation since we use all principal components.
More explicitly, denote $X_{1:N}$ as the matrix with the first $N$ states of the network data represented by row vectors. If $X_{1:N}=U\Sigma P^T$ here represents the singular value decomposition ($U$ and $V$ are unitary matrices and $\Sigma$ is a diagonal matrix) of $X_{1:N}$ and we refer to \cite{wikiSVdecomp}. We then denote the principal component matrix $P$, and the principal components are given by 
 $Z_{1:N}=X_{1:N}P$.
These principal components are then used to train a feedforward neural network. Denote the row vectors of $Z_{1:N}$ by 
$z_i^T, i=1,2,\ldots,N$ and the neural network by $NN$, the network is subsequently trained to learn an approximation of the map
$NN: (z_{n-1},z_n) \mapsto  u_n.
$
The learnt approximation of $NN$ can be used to approximate $\Gamma$ since 
\begin{equation}\label{Seq_RNN}
  NN\left( \begin{bmatrix} 
    P^Tx_{n-1} \\
    P^Tx_n
    \end{bmatrix}
    \right) = NN \circ 
    \begin{bmatrix}
    P^T & 0 \\
    0 & P^T 
    \end{bmatrix}\begin{bmatrix}
    x_{n-1}\\
    x_n
    \end{bmatrix} = u_n = \Gamma(x_{n-1},x_n).
\end{equation}

\section{Experimental Results}
In this section we showcase numerical results pertaining to 3 attractors that were simulated using the theory and methodology described. We first consider a simple physical system that exhibits complicated dynamics.
\subsection{Double Pendulum}

The Double Pendulum consists of two pendulums fastened to one another such that the system moves as a whole. The first (or higher pendulum) is attached to a fixed point while the second is attached to the endpoint of the first. (See below the Figure~\ref{fig:dp_setup})
This system shows chaos strikingly. Consider again Figure~\ref{fig:dp_sdic}.
The system has four variables: angle and angular velocity of each pendulum rod. An ideal system is assumed where the rods are massless and no friction in the pivots are considered.
\begin{figure}[ht]
  \includegraphics[scale=0.35]{Graphs/_dp_setup.eps}
  \centering
  \captionof{figure}{Pictorial representation of the setup of a DP. Image taken from~\cite{DPSetup}}
\label{fig:dp_setup}
\end{figure}

The scalar equations of motion are derived using Newtonian physics and are reproduced from~\cite{DPFormulas} in Equations~\ref{eqns_dp}. Each pendulum component has an associated mass of the head, $m$, and length of the rod, $L_i$. Acceleration due to gravity is denoted by $g$ and initial velocities by $\dot{\theta}$
\begin{eqnarray}\label{eqns_dp}
  \ddot{\theta_{1}}  = \frac{-g(2m_1+m_2)\sin{\theta_1} - m_2g\sin(\theta_1-2\theta_2) - 2\sin(\theta_1-\theta_2)m_2({\dot{\theta_{2}}}^{2}L_2 + {\theta_{1}'}L_1\cos(\theta_1-\theta_2))} {L_1(2m_1 + m_2 -m_2\cos(2\theta_1 - 2\theta_2))}
  \\
  \ddot{\theta_{2}} = \frac{2\sin(\theta_1-\theta_2)(\theta_{1}'^{2}L_1(m_1+m_2) + g(m_1+m_2)\cos\theta_1 + \dot{\theta_{2}}^{2}L_2m_2\cos(\theta_1-\theta_2))}{L_2(2m_1 + m_2 -m_2\cos(2\theta_1 - 2\theta_2))}
\end{eqnarray}

% The second-order system is converted to a first-order ODE system and solved using the Runge-Kutta numerical method.  
Here we chose the values $m_1=2.0$, $m_2=2.0$, $L_1=1.5$ and $m_2=1.5$ with initial velocities $v_1=1.5, v_2=-2$ for the respective pendulum heads.
With a change of variables the second-order system can be 
written as a system of first order equations with four variables, where two variables devoted to each pendulum denote the $x$ and $y$ coordinates of the mass (the lengthy equations are not shown here). The ODE system is then solved using the fourth-order Runge-Kutta numerical method.

% Undamped DP data-sets were generated and used in this project due to the requirement that the map $T$ be surjective that we must not forget. 
% An example of this would be the case of a DP exhibiting energy-loss (i.e. where the map $T$ is not surjective). In such a case, the attractor is a single point in the phase space when the double pendulum comes to rest. 
% It must be mentioned once again that if we use data outside the attractor, learning is not reliable. 
% To see this, consider Figure~\ref{fig:damped_pendulum} and~\ref{fig:dp_notsurjective}.

We need the assumption the system is not damped so that the map $T$ resulting from the discretization of the system  is surjective.  If we assume damping surjectivity is lost on any phase space with more than one point.  
An example of this would be the case of a DP exhibiting energy-loss. In such a case, the attractor is a single point in the phase space when the double pendulum comes to rest. 
For such a system, it may be noted that if we use data outside the attractor, learning is not reliable. To see this, consider the plots ~\ref{fig:damped_pendulum} and ~\ref{fig:dp_notsurjective}.

@@

\begin{figure}[ht]
  \includegraphics[scale=0.3]{Graphs/_dpfail_nonsurj.eps}
  \centering
\caption{Failure to forecast or learn a map which is not surjective: Evolution of the $y$-coordinate of the bottommost pendulum plotted here for 15000 timesteps. Plotted here the predicted value (red) versus the actual value (blue). The plots overlap for the first 10000 steps, the duration of training to learn the map $\Gamma$. Parameter-values are exactly as in the successful case discussed in the paragraph directly afterwards. Afterwards, the trajectories start diverging and it the predicted value no longer accurately represents the complexity or energy-loss of the actual data. }
\label{fig:dp_notsurjective}
\end{figure}
% \ednote{Can I say in the caption \emph{Parameter-values are exactly as in the successful case discussed in the paragraph directly afterwards.}}

@@

The system was driven for 15500 steps, of which the first 500 were immediately discarded to simulate the network `forgetting'. The driving function $g$ had input and reservoir matrices were implemented as sparse matrices filled to 20\%. Parameter-values of $\alpha=0.99$ and $a=0.5$ were used.
The remaining 15000 datapoints were used to train a network with 128 neurons. 
The network was initialised to contain 16 hidden layers each with a layer dimension of 64, activated with the $\tanh$ activation function.
Training was accomplished using 256 epochs with batch sizes 128 and a delay of 1. See Fig.~\ref{fig:dp_success_traj} for results pertaining to the trajectory taken by the DP's components and then Fig.~\ref{fig:dp_success_density} to witness associated densities of predicted versus actual values.

@@

We now consider an experiment where noise is added to the system and find it to be incredibly robust. We employ the same parameters as in the preceding experiment and add noise from a normal-distribution with mean zero and standard deviation equal to 0.1 which gives a signal-to-noise ratio of 39dB.

For the signal-to-noise ration we adopt the formula $\text{SNR}_{dB}=10\text{log}_{10}\big(\frac{P_\text{{signal}}}{P_{\text{noise}}}\big)$ where $P_\text{{signal}}$ refers to the variance observed in the normalised data and $P_{\text{noise}}$ represents the variance of the added noise. \ednote{B:Added this sentence.}
Once again consider the actual and predicted trajectories for the noisy DP dataset in Figure~\ref{fig:noisydp_success_traj}. Next we consider the densities of the predicted and actual values in Figure~\ref{fig:noisydp_success_density} and observe that they display highly similar distributions.

@@

%  Lastly, a run was done where only a single coordinate (the $y$-coordinate of the lower pendulum head) was fed into the system and we can illustrate through the obtained success that the coordinate contains all the information in its past states necessary to forecast its future evolution.
%  \textbf{Insert graphs here. Had to quickly change something here.}

@@

The behaviour of the system with a $\beta$-parameter value of 0.1056 was simulated by scaling the data from it to fit inside the interval $[-1,1]$. Two cases were considered: one noise-free and another perturbed by noise. 
Noise was generated  by adding data points taken from a normal distribution with mean 0 and standard deviation 0.05, which translates approximately to a signal-to-noise ration of 33dB. 
($\text{SNR}_{dB}$ is calculated exactly as for the DP-data above)

@@

Another two-equation discrete-time map named the Fractal Dream Attractor, or more commonly as the Pickover or Clifford Map (first discovered by Clifford A. Pickover and discussed in his fascinating book``Chaos in Wonderland" \cite{PickoverChaos}), was also considered.
\begin{eqnarray}\label{eqns_clifford}
  {x_{n+1}=\sin(ay_n) + c\cos(ax_n)} \\
  {y_{n+1}=\sin(bx_n)+d\cos(by_n)}
\end{eqnarray}

@@

The behaviour of the system with a parameter values of $x_0=0$, $y_0=0$, $a = -1.7$,  $b = 1.8$, $c = -1.9$ and $d = -0.4$ was examined. 
Data was scaled to fit inside the interval $[-1,1]$. Only the noise-free case was investigated. 
Depictions of the empirical results available in Figure~\ref{fig:Clifford}.
 A sequence of 25000 observations was fed into the system with 15000 entries discarded \ednote{M: Why did you have to discard 15000?} to simulate the network's loss of memory. The next 1000 steps were predicted and compared with true data.

 The RNN was initialised exactly as for Thomas' Cyclically Symmetric Strange attractor: the dimension of the network is set equal to 500 neurons and 12 hidden layers of dimension 64 each were initialised. Training is realised through 150 epochs of batch sizes of 128 each; the map is learnt via the Adam Optimiser using finer and finer learning rates (0.001 and then divided by 10 at every iteration).

\section{Functional Complexity Reduction in Noninvertible Maps}
% Besides being able to learn the action of the Koopman operator exactly, our setup has other advantages over selecting or optimizing the choice of observables as in the Extended Dynamic Mode Decomposition (EDMD) algorithm (see, [21, 22]). We can alter the functional complexity of G T by changing the number of observables or equivalently by altering the dimension of X – in the case of a RNN implementation of g, by simply changing the number of neurons in the network. Empirically, increasing the dimension of X is found to increase the linear relationship (or intuitively reduces the functional complexity of G T ) that is measured as a generalization of the Pearson correlation coefficient to random vectors (see, [44]) between (x_{n−1} , x_{n}) and G_T (x_{n−1} , x_n), and such numerical evidence is tabulated in [11, Table 1]. In contrast, expanding the set of observables while using EDMD to learn a map with a lower functional complexity is a highly involved task, not least because so much that one often does not know how to guess a new observable, as well as the fact that adding observables does not necessarily retain the invariance of the span of the observables.
